Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population

### 2023-09-12 analysis/paper/paper.tex
add R package check workflow

\usepackage{lineno}
\linenumbers
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population},
  pdfauthor={Bjørn Peare Bartholdy1,; Jørgen B. Hasselstrøm2; Lambert K. Sørensen2; Maia Casna1; Menno Hoogland1; Historisch Genootschap Beemster3; Amanda G. Henry1},
  pdfkeywords={dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Multiproxy analysis exploring patterns of diet and disease in
dental calculus and skeletal remains from a 19th century Dutch
population}
\author{}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Dental calculus is an excellent source of information on the dietary
patterns of past populations, including consumption of plant-based
items. The detection of plant-derived residues such as alkaloids and
their metabolites in dental calculus provides direct evidence of
consumption by individuals within a population. We conducted a study on
41 individuals from Middenbeemster, a 19th century rural Dutch
archaeological site. Skeletal and dental analysis was performed to
explore potential relationships between pathological conditions/lesions
and the presence of alkaloids. We also explored other factors
potentially affecting the detection of alkaloids, including sample
weight and skeletal preservation. Dental calculus was sampled and
analysed using ultra-high-performance liquid chromatography-tandem mass
spectrometry (UHPLC-ESI-MS/MS). We were able to detect nicotine,
cotinine, caffeine, theophylline, and salicylic acid. By detecting these
compounds we are able to show the consumption of tea and coffee and
smoking of tobacco on an individual scale, which is also confirmed by
historic documentation and identification of pipe notches in the
dentition. Nicotine and/or cotinine was present in 60\% of individuals
with at least one visible pipe notch. We find some influence of skeletal
preservation on the detection of alkaloids and salicylic acid, with
higher quantities of compounds extracted from well-preserved
individuals, and also observe a relationship between weight of the
calculus sample and raw quantity of the detected compounds, and we were
able to detect alkaloids in samples as small as 2 mg. We found
correlations between chronic maxillary sinusitis and the presence of
multiple alkaloids. We show that there are many limitations that will
need to be addressed going forward with this type of analysis, and
stress the need for more systematic research on the consumption of
alkaloid-containing items and their subsequent concentration and
preservation in dental calculus, in addition to how mode of consumption
may affect concentrations on different parts of the dentition. Despite
the limitations, this preliminary study illustrates the many benefits of
using calculus to target a variety of compounds that could have been
ingested as medicine or diet, or consumed in a different manner. This
method allows us to directly address specific individuals, which can be
especially useful in individuals that are not always well-documented in
historic documentation, such as rural populations, children and women.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, boxrule=0pt, sharp corners, frame hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

\textsuperscript{1} Department of Archaeological Sciences, Leiden
University\\
\textsuperscript{2} Department of Forensic Medicine, Aarhus University\\
\textsuperscript{3} Historisch Genootschap Beemster

\textsuperscript{*} Correspondence:
\href{mailto:b.p.bartholdy@arch.leidenuniv.nl}{Bjørn Peare Bartholdy
\textless{}b.p.bartholdy@arch.leidenuniv.nl\textgreater{}}

Keywords: dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco

>>>>

\usepackage{lineno}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population},
  pdfauthor={Bjørn Peare Bartholdy; Jørgen B. Hasselstrøm; Lambert K. Sørensen; Maia Casna; Menno Hoogland; Historisch Genootschap Beemster; Amanda G. Henry},
  pdfkeywords={dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Multiproxy analysis exploring patterns of diet and disease in
dental calculus and skeletal remains from a 19th century Dutch
population}
\author{Bjørn Peare Bartholdy \and Jørgen B. Hasselstrøm \and Lambert K.
Sørensen \and Maia Casna \and Menno Hoogland \and Historisch Genootschap
Beemster \and Amanda G. Henry}
\date{}

\begin{document}
%% modified from: https://github.com/quarto-journals/plos/blob/main/_extensions/plos/partials/before-body.tex
\begin{flushleft}
{\LARGE
\textbf\newline{Multiproxy analysis exploring patterns of diet and
disease in dental calculus and skeletal remains from a 19th century
Dutch population}
}
\newline

\vspace{12pt}
% Insert author names, affiliations and corresponding author email

\normalsize{Bjørn Peare
Bartholdy}\textsuperscript{1*}, \normalsize{Jørgen B.
Hasselstrøm}\textsuperscript{2}, \normalsize{Lambert K.
Sørensen}\textsuperscript{2}, \normalsize{Maia
Casna}\textsuperscript{1}, \normalsize{Menno
Hoogland}\textsuperscript{1}, \normalsize{Historisch Genootschap
Beemster}\textsuperscript{3}, \normalsize{Amanda G.
Henry}\textsuperscript{1}
\\
\bigskip
\small{1} \small{Department of Archaeological Sciences, Leiden
University}\\ \small{2} \small{Department of Forensic Medicine, Aarhus
University}\\ \small{3} \small{Historisch Genootschap Beemster}
\newline

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* b.p.bartholdy@arch.leidenuniv.nl

\end{flushleft}

\linenumbers

\begin{abstract}
Dental calculus is an excellent source of information on the dietary
patterns of past populations, including consumption of plant-based
items. The detection of plant-derived residues such as alkaloids and
their metabolites in dental calculus provides direct evidence of
consumption by individuals within a population. We conducted a study on
41 individuals from Middenbeemster, a 19th century rural Dutch
archaeological site. Skeletal and dental analysis was performed to
explore potential relationships between pathological conditions/lesions
and the presence of alkaloids. We also explored other factors
potentially affecting the detection of alkaloids, including sample
weight and skeletal preservation. Dental calculus was sampled and
analysed using ultra-high-performance liquid chromatography-tandem mass
spectrometry (UHPLC-ESI-MS/MS). We were able to detect nicotine,
cotinine, caffeine, theophylline, and salicylic acid. By detecting these
compounds we are able to show the consumption of tea and coffee and
smoking of tobacco on an individual scale, which is also confirmed by
historic documentation and identification of pipe notches in the
dentition. Nicotine and/or cotinine was present in 60\% of individuals
with at least one visible pipe notch. We find some influence of skeletal
preservation on the detection of alkaloids and salicylic acid, with
higher quantities of compounds extracted from well-preserved
individuals, and also observe a relationship between weight of the
calculus sample and raw quantity of the detected compounds, and we were
able to detect alkaloids in samples as small as 2 mg. We found
correlations between chronic maxillary sinusitis and the presence of
multiple alkaloids. We show that there are many limitations that will
need to be addressed going forward with this type of analysis, and
stress the need for more systematic research on the consumption of
alkaloid-containing items and their subsequent concentration and
preservation in dental calculus, in addition to how mode of consumption
may affect concentrations on different parts of the dentition. Despite
the limitations, this preliminary study illustrates the many benefits of
using calculus to target a variety of compounds that could have been
ingested as medicine or diet, or consumed in a different manner. This
method allows us to directly address specific individuals, which can be
especially useful in individuals that are not always well-documented in
historic documentation, such as rural populations, children and women.
\end{abstract}

Keywords:  dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco 
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, frame hidden, breakable, enhanced]}{\end{tcolorbox}}\fi
